Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-26 manuscript.qmd
reword first part of results

In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps. Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors. The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%. Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo). The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-algebra-equation, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrices represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrices describe the evolution of the contralateral hidden states once for the case of no midline extension (top) and once for the case where the tumor does extend over the mid-sagittal line (bottom). In the center, the time-prior for late T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for late T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure @fig-model-state-dist. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV for a late T-category tumor. Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps. These two matrices are the result of @eq-bilateral-marginal-algebra, which we visualized as well in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

The bilateral model was designed to fulfil the requirements laid out in @sec-requirements. Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset. To that end, we compare the model's predictions for contralateral involvement against observations in the data. This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.

@@

In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.

@@

@Fig-model-prevalences-overall shows that the model is capable of accurately accounting for the most important risk factors, i.e. T-category and midline extension. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line. Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.

@@

![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-ctageory (early or advanced). Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}

@Fig-model-prevalences-upstream illustrates how rarely the contralateral LNL III harbors metastases when its upstream level II is healthy. Also, the model can capture this correlation rather well, with the exception of early T-category tumors that extend over the mid-sagittal line. As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.

@@

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. Shown are marginals of the joint distribution where we consider contralateral LNL II involvement with simulaneous involvement of different ipsilateral LNLs. The model has no direct connections between the two sides and thus no adjustable parameter to specifically quantify correlations between ipsi- and contralateral involvement. Metastases in the two sides of the neck are only correlated via the time of diagnosis. Nevertheless, the model adequately predicts that contralateral level II involvement is rare in combination with a healthy ipsilateral neck (green histograms). If ipsilateral LNL II is involved, contralateral involvement is more frequent.

>>>>

In this subsection, we illustrate key aspects of the mathematical framework introduced earlier. The top panel of @fig-model-midext-evo shows the prior distribution over diagnosis times, $P(t)$. Based on the parameterization, early T-category tumors are on average diagnosed after {{< var model.early_expected_time >}} time steps, while advanced T-category tumors are diagnosed later, averaging {{< var model.late_expected_time >}} time steps. This is due to the learned value of $p_\text{adv.}$, which is {{< var model.params_percent.late_p >}}% for advanced T-category tumors. The tumor's average probability per time step of growing over the midline, $p_\epsilon$, was found to be {{< var model.params_percent.midext_prob >}}%. Using this value, the conditional probability of midline extension, $P(\epsilon \mid t)$, can be computed for a given time step $t$ (red line in the top panel of @fig-model-midext-evo). The bottom panel visualizes the joint probability $P(\epsilon, t)$, showing the likelihood of diagnosis at time $t$ with specific states of midline extension and T-category.

![The top panel shows the prior probability to be diagnosed at time step $t$ for early and late T-category tumors as bars. The conditional probability of midline extension ($\epsilon=\texttt{True}$) given time step $t$ is shown as a line plot. The bottom panel illustrates the joint probability of being diagnosed at time $t$ and having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}

The framework models the joint probability distribution of midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-bilateral-marginal-algebra. To simplify interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 16$ possible states per side, and $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII are excluded, along with their spread parameters, while remaining parameters are set to their mean values from @tbl-midline-params.

![Visual representation of @eq-bilateral-marginal-algebra for midline extension. The left and right matrices represent the time evolution of hidden states for the ipsi- and contralateral necks, respectively. The right matrices distinguish between the cases of no midline extension (top) and midline extension (bottom). The central diagonal matrix shows the time-prior for late T-category tumors. This computation yields the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist.](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

The left matrix in @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement states, starting from the healthy state $[0,0,0]$. The two right matrices show the contralateral state evolution, distinguishing between the cases of midline extension and no midline extension. At $t=0$, the contralateral neck begins in the healthy state $[0,0,0]$ without midline extension. The central matrix shows the time prior for late T-category tumors. The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$, visualized in @fig-model-state-dist.

This joint distribution is presented as two heatmaps, corresponding to the two states of midline extension. The most likely state involves a lateralized tumor with ipsilateral level II involvement and no contralateral involvement, having a probability of approximately 25%. The next most probable state involves ipsilateral levels II and III without contralateral involvement. The most likely state with contralateral involvement corresponds to tumors with midline extension, showing involvement of contralateral level II and ipsilateral levels II and III.

![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over ipsi- and contralateral states and midline extension for late T-category tumors. This three-dimensional distribution is shown as two separate heatmaps for the binary midline extension variable $\epsilon$. These matrices are the result of @eq-bilateral-marginal-algebra, visualized in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

The bilateral model was designed to meet the requirements outlined in @sec-requirements. Here, we evaluate the model's ability to quantitatively capture the observed patterns of lymph node involvement in the dataset. Specifically, we compare the model's predictions for contralateral involvement to the observed data across scenarios that vary by T-category, midline extension, and ipsilateral involvement.

@@

In @fig-model-prevalences-overall, we compare the prevalence of contralateral involvement for LNLs II, III, and IV across four scenarios: combinations of early and late T-category, and lateralized versus midline-extending tumors.

@@

@Fig-model-prevalences-overall demonstrates the model's ability to account for key risk factors, such as T-category and midline extension. Consistent with the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% for advanced T-category tumors that cross the midline. Similarly, contralateral LNL III involvement rises from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to nearly {{< var prevalence.late_ext_iN0_cIII.predicted >}}% for advanced T-category midline-extending tumors.

@@

![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-category (early or advanced). Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}

@Fig-model-prevalences-upstream highlights the influence of upstream LNL II involvement on contralateral LNL III metastasis. The contralateral LNL III rarely harbors metastases when its upstream LNL II is healthy, a correlation well-captured by the model. However, for early T-category tumors extending over the midline, the model slightly underestimates the observed prevalence, likely due to the relative rarity of such cases, resulting in broader uncertainty in the observed prevalence estimates.

@@

In @fig-model-prevalences-with-ipsi, the model's ability to capture the correlation between ipsi- and contralateral involvement is demonstrated. The marginals of the joint distribution highlight contralateral LNL II involvement alongside varying ipsilateral LNL involvement states. Despite having no direct connections between the two sides, the model successfully predicts these correlations, which arise purely through the shared diagnosis time.

For example, the model accurately predicts that contralateral LNL II involvement is rare when the ipsilateral neck is completely healthy (green histograms). However, if ipsilateral LNL II is involved, contralateral involvement becomes more likely. Notably, the model achieves this without specific parameters to quantify ipsi-contralateral correlations, relying instead on the inherent structure of the time-dependent dynamics.
